---
title: "Exercise 03"
author: "Wannisa Ritmahan"
date: today
format: 
    html:
      toc: true
      self-contained: true
      highlight-style: github
      code-line-numbers: true
editor_options: 
    chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load Packages and Data

Use the Rsubread package to map a read file found at http://fgcz-gstore.uzh.ch/public/Yeast_Reads/


```{r, eval=FALSE}
BiocManager::install("Rsubread", update=FALSE)
BiocManager::install("BSgenome.Scerevisiae.UCSC.sacCer3", update=FALSE)

# download the read data file
download.file("http://fgcz-gstore.uzh.ch/public/Yeast_Reads/20170313.A-G1_R1-sub.fastq.gz", "G1_R1-sub.fastq.gz")


```



```{r, message=FALSE}
library(BSgenome)
library(BSgenome.Scerevisiae.UCSC.sacCer3)
library(Rsubread)
library(dplyr)
library(GenomicAlignments)
library(formattable)
```

Use the commands *export* to export method in the *BSgenome* package to write the genome as a fasta file on the local disk. Use the commands *buildindex* and *align* of the *Rsubread* package to map the reads. Consult the help pages of the respective functions. Build the index so that at most 2GB of RAM is used. Run the alignment with 4 threads. If you computer has less resources, adapt accordingly.

We get the reference sequence from the genome package *BSgenome.Scerevisiae.UCSC.sacCer3* and we export it as a fasta file.

```{r, eval=FALSE}
# Fetch reference genome (fasta) and write on the local disk
export(BSgenome.Scerevisiae.UCSC.sacCer3, con="sacCer3-genome.fasta")

# Build a base-space index (default)
buildindex("sacCer3-Rsub", "sacCer3-genome.fasta", memory=2000 )
```

## Align

Align to the genome with the *RSubread::align* function that does not explicitely look for spliced alignements. Do the alignment for the full length reads. 
Then run the alignment also a second time with fixed trimming of 10 bases from both ends, i.e. use *align(....., nTrim5 = 10, nTrim3 = 10)*.

```{r}

# ------------------------------------------------------------
#  Align full read length; Phred 33 (from fastQC report)
# ------------------------------------------------------------
alignFull.stat <- align(index = "sacCer3-Rsub",
                        readfile1="G1_R1-sub.fastq.gz",
                        output_file="G1_R1-sub.bam", 
                        type = "rna",
                        phredOffset = 33, 
                        nthreads = 4)

# ------------------------------------------------------------
# Align with options to trim 10 nt from both ends of the reads
# ------------------------------------------------------------
alignTrim.stat <- align(index = "sacCer3-Rsub",
                        readfile1="G1_R1-sub.fastq.gz",
                        output_file="G1_R1-sub_trim10.bam", 
                        type = "rna",
                        phredOffset = 33, 
                        nthreads = 4,
                        nTrim5 = 10, 
                        nTrim3 = 10)

# ------------------------------------------------------------
# Summary alignment stats
# ------------------------------------------------------------
align_stats <- merge(alignFull.stat, 
                    alignTrim.stat, 
                    by = 'row.names', all = TRUE) 
colnames(align_stats) <- c("Alignment Stats", "Default", "Trimmed")
align_stats["Difference"] <- align_stats["Trimmed"] - align_stats["Default"]
formattable(align_stats)

```

## Question 1: For both runs (with and without trimming), what fraction of the reads aligns uniquely and multiple times?

From untrimmed alignment, 2,408,263 from total reads of 2,667,280 (90.3%) were mapped to the reference. 80.1% (1929371/2408263) of mapped reads were uniquely mapped while 19.9 % (478892/2408263) mapped to multiple locations.

Trimming 10 nucleotides from both sides of each read improved the number of mapped reads to 2,483,097 (93.1%). The proportion of unique mapped reads is however decreased to 79.4% (1972065/2483097), inevitably increases multiple mapped reads to 20.6% (511032/2483097).

## Question 2: Are the reads that align the same in both cases?

Check the agreement of the alignments. Load the uniquely aligning reads. Use the function *readGAlignments*.
Use the parameter settings to read also the read id, and the number of hits (NH)

Hint:
```{r}

# ------------------------------------------------------------
# Define BAM tags (NH = Number of hits) in BAM files
# ------------------------------------------------------------
sbp <- ScanBamParam(what = c("qname"), tag="NH")

# ------------------------------------------------------------
# Scan BAM alignment files with defined search parameters
# ------------------------------------------------------------
alignFull <- readGAlignments('G1_R1-sub.bam', param=sbp)
alignTrim <- readGAlignments('G1_R1-sub_trim10.bam', param=sbp)

# ------------------------------------------------------------
# Find common read ID mapped to the reference
# Note: mcols(alignFull) has two columns; qname and NH
# ------------------------------------------------------------
mapped_ids = intersect(mcols(alignFull)$qname, 
                       mcols(alignTrim)$qname)
print(paste0("Common mapped reads: ", length(mapped_ids)))

```

Alternatively, flag `scanBamFlag(isUnmappedQuery = FALSE)` can be used.

```{r}

# ------------------------------------------------------------
# Extract only mapped read IDs
# ------------------------------------------------------------
sbp_mapped <- ScanBamParam(
  flag = scanBamFlag(isUnmappedQuery = FALSE), what = "qname")

# ------------------------------------------------------------
# Look up for unmapped reads
# ------------------------------------------------------------
alignFull_m <- scanBam('G1_R1-sub.bam', param = sbp_mapped)
alignTrim_m <- scanBam('G1_R1-sub_trim10.bam', param = sbp_mapped)

# ------------------------------------------------------------
mapped_ids2 <- intersect(unique(alignTrim_m[[1]]$qname), 
                          unique(alignFull_m[[1]]$qname))

print(paste0("Common mapped reads: ", length(mapped_ids2)))

```

## Question 3: For those reads that align full and trimmed, do they align to the same location?

Hint:

```{r }

# ------------------------------------------------------------
# Include only common mapped ids for comparisons
# ------------------------------------------------------------
alignFull_mc <- alignFull[match(mapped_ids, mcols(alignFull)$qname)]
alignTrim_mc <- alignTrim[match(mapped_ids, mcols(alignTrim)$qname)]

# ------------------------------------------------------------
# Finding overlaps of two matching sets element-by-element 
# ------------------------------------------------------------
ovlStatus <- poverlaps(granges(alignTrim_mc), 
                       granges(alignFull_mc), 
                       type="within") %>% as.vector()

print(paste0("Common mapped reads that align at the same position: ", length(ovlStatus), " (", round(mean(ovlStatus)*100, digits = 2), "%)"))

```
Most of commonly mapped reads between full and trimmed reads aligned at the same locations.

## `sessionInfo()`

```{r}
sessionInfo()
```


## AI declaration

No AI agents/tools used for creating code or report.

